Northern Pikeminnow Predation on Juvenile Chinook Salmon and Potential Impacts to Local Populations in the Upper Salmon River, Idaho

### 2021-12-16 analysis/paper/paper.Rmd
minor edits from Google Doc

The Deadwater Slough is in a reach of the Salmon River believed to be a historically important overwinter rearing and spring emigration area for Chinook Salmon juveniles. Spring/summer Chinook Salmon in the Upper Salmon MPG are stream-type and exhibit two distinct migration tactics; downstream rearing (DSR) and natal reach rearing (NRR) [@Copeland2014a]. The DSR migrants leave the natal spawning area as subyearlings between June and november and typically overwinter in downstream, mainstem habitats until the following spring when they emigrate to the ocean as smolts. Alternatively, NRR migrants remain in their natal spawning areas for approximately one year after emergence until emigration to the ocean as smolts. Diversity of migratory tactics provides a mechanism for coping with adverse conditions in freshwater rearing and migration environments and buffers against catastrophic events, thereby increasing population resiliency. Because the Deadwater Slough both supports rearing for DSR migrants and provides the migration pathway for NRR migrants, it represents a significant habitat for the Upper Salmon MPG.

@@

## Abundance of Piscivorous Fishes

@@

To estimate the total consumption potential (i.e., number/amount of juvenile consumed during the peaks of DSR [fall] and NRR [spring] emigrations) of Northern Pikeminnow in Deadwater Slough, we used the Fish Bioenergetics v4.0 application developed by @Deslauriers2017 applied in the R statistical software [@RCoreTeam2021]. The daily rate of consumption in grams for an individual Northern Pikeminnow was estimated based on predator and prey energy densities, predator start and end weights, and water temperatures. Predator energy density for Northern Pikeminnow was fixed at 6,703 Joules(J)/g [@Deslauriers2017]. Prey energy densities were fixed at 21,500 J/g for juvenile Chinook Salmon based on an estimate from @Moss2016. We used a weight-length formula [@Parker1995] to calculate the average predator start weight from the average length of Northern Pikeminnow captured during our study using the FSA package in R [@Ogle2021]. The average length of Northern Pikeminnow captured at the Deadwater Slough during the spring and fall were 352.9 and 393.7 mm, respectively, which calculates to average starting weights of 504.4 and 670.5 g. Continuous water temperature data from 2019 was used in two alternative models, one using the baseline start and end weights and the second with a 10% increase in average end weight on Northern Pikeminnow.

@@

To estimate the total biomass consumed, we then multiplied the biomass consumed by an individual Northern Pikeminnow for the given time period by the estimated population size for that time period. The total biomass was further converted to an estimate of the total number of Chinook Salmon consumed by dividing by a weight of 12 g per individual. This weight was an average from juvenile Chinook Salmon emigrating past six rotary screw traps operating in the Upper Salmon above Deadwater Slough including traps in the Lemhi, Pahsimeroi, and North Fork Salmon rivers and one trap operating near the Sawtooth hatchery.

@@

We estimated the population size of Northern Pikeminnow in the Deadwater Slough to be greater than ***xx,xxx*** during the fall emigration period for DSR Chinook Salmon. That estimate translates to a density of ***xxx*** Northern Pikeminnow per 100 m or ***xxx*** per 100 m^2^ which is ***similar/more/less*** than estimates from elsewhere in the Columbia River ***(citation)*** where substantial Northern Pikeminnow predation impacts on salmonids have led to bounty programs aimed at reducing Northern Pikeminno abundance. The population size of Northern Pikeminnow was not directly estimated during the spring NRR Chinook salmon emigration period however, the relative abundance measured at CPUE was comparable to the fall sampling periods ***(update statement later)***. The population of Northern Pikeminnow in Deadwater Slough was estimated to consume between ***xx,xxx*** and ***xx,xxx*** juvenile Chinook Salmon during the ***x*** sampling periods and result in an estimated reduction of returning adults between ***xxx*** and ***x,xxx***. We suggest that the habitat modifications that created the Deadwater Slough have resulted in favorable conditions for Northern Pikeminnow, including improved conditions for predation upon juvenile Chinook Salmon ***(add detail here)***. Therefore, predation by Northern Pikeminnow in the Deadwater Slough likely has a consequential impact on ESA-listed Chinook Salmon populations in the Upper Salmon River MPG.

@@

The majority of fish collected during this study received gastric lavage, including some non-predatory species. To validate the efficacy of this method, we euthanized nine Northern Pikeminnow after gastric lavage was completed and removed the remaining stomach contents via dissection. Our results support previous findings that gastric lavage effectively removes stomach contents (@Lott2020). Additionally, we found that the fish captured in net-traps had a similar proportion of stomach content samples. of those, 1,214 (76.2%) were completely empty, 345 (21.6%) had stomach contents, and 35 (2.2%) contained fish remnants. We were successfully able to identify Redside Shiners, a Largescale Sucker, sculpin, Mountain Whitefish, and one juvenile Chinook Salmon in the stomach contents. However, decomposition from digestion rendered most stomach contents unidentifiable.

@@

The authors extent much appreciation to the many volunteers who assisted with field efforts including collaborators from Bureau of Reclamation, Idaho Department of Fish and Game, and Lemhi Regional Land Trust, among others. This manuscript benefited from reviews and contributions from colleagues at the Idaho Governor's Office of Species Conservation, Rio Applied Science and Engineering, and from Sean Gibbs and Ben Briscoe at Mount Hood Environmental. Funding for this study was provided by the Bureau of Reclamation, Pacific Northwest Regional Office (contract No. 140R1021F0018). Special thanks to Caitlin Alcott and Inter-Fluve for their administrative support and guidance.

>>>>

The Deadwater Slough is in a reach of the Salmon River believed to be a historically important overwinter rearing and spring emigration area for Chinook Salmon juveniles. Spring/summer Chinook Salmon in the Upper Salmon MPG are stream-type and exhibit two distinct migration tactics; downstream rearing (DSR) and natal reach rearing (NRR) [@Copeland2014a]. The DSR migrants leave the natal spawning area as subyearlings between June and November and typically overwinter in downstream, mainstem habitats until the following spring when they emigrate to the ocean as smolts. Alternatively, NRR migrants remain in their natal spawning areas for approximately one year after emergence until emigration to the ocean as smolts. Diversity of migratory tactics provides a mechanism for coping with adverse conditions in freshwater rearing and migration environments and buffers against catastrophic events, thereby increasing population resiliency. Because the Deadwater Slough both supports rearing for DSR migrants and provides the migration pathway for NRR migrants, it represents a significant habitat for the Upper Salmon MPG.

@@

## Northern Pikeminnow Abundance

@@

To estimate the total consumption potential (i.e., number/amount of juvenile consumed during the peaks of DSR [fall] and NRR [spring] emigrations) of Northern Pikeminnow in Deadwater Slough, we used the Fish Bioenergetics v4.0 application developed by @Deslauriers2017 applied in the R statistical software [@RCoreTeam2021]. The daily rate of consumption in grams for an individual Northern Pikeminnow was estimated based on predator and prey energy densities, predator start and end weights, and water temperatures. Predator energy density for Northern Pikeminnow was fixed at 6,703 Joules(J)/g [@Deslauriers2017]. Prey energy densities were fixed at 21,500 J/g for juvenile Chinook Salmon based on an estimate from @Moss2016 and 3,000 J/g for invertebrates. Additional settings included 0% assumed indigestible prety, and a variation in the proportion of diet that was invertebrates or fish ranging from 10% to 90% for both categories. We used a weight-length formula [@Parker1995] to calculate the average predator start weight from the average length of Northern Pikeminnow captured during our study using the FSA package in R [@Ogle2021]. The average length of Northern Pikeminnow captured at the Deadwater Slough during the spring and fall were 352.9 and 393.7 mm, respectively, which calculates to average starting weights of 504.4 and 670.5 g. Continuous water temperature data from 2019 was used in two alternative models, one using the baseline start and end weights and the second with a 10% increase in average end weight on Northern Pikeminnow.

@@

To estimate the total biomass consumed, we then multiplied the biomass consumed by an individual Northern Pikeminnow for the given time period by the estimated population size for that time period. The total biomass was further converted to an estimate of the total number of Chinook Salmon consumed by dividing by a weight of 12 g per individual. This weight was an average from juvenile Chinook Salmon captured at six rotary screw traps throughout the operation season March to November in the Upper Salmon above Deadwater Slough including traps in the Lemhi, Pahsimeroi, and North Fork Salmon rivers and one trap operating near the Sawtooth hatchery.

@@

We estimated the population size of Northern Pikeminnow in the Deadwater Slough to be greater than ***xx,xxx*** during the fall emigration period for DSR Chinook Salmon. That estimate translates to a density of ***xxx*** Northern Pikeminnow per 100 m or ***xxx*** per 100 m^2^ which is ***similar/more/less*** than estimates from elsewhere in the Columbia River ***(citation)*** where substantial Northern Pikeminnow predation impacts on salmonids have led to bounty programs aimed at reducing Northern Pikeminnow abundance. The population size of Northern Pikeminnow was not directly estimated during the spring NRR Chinook salmon emigration period; however, the relative abundance measured at CPUE was comparable to the fall sampling periods ***(update statement later)***. The population of Northern Pikeminnow in Deadwater Slough was estimated to consume between ***xx,xxx*** and ***xx,xxx*** juvenile Chinook Salmon during the ***x*** sampling periods and result in an estimated reduction of returning adults between ***xxx*** and ***x,xxx***. We suggest that the habitat modifications that created the Deadwater Slough have resulted in favorable conditions for Northern Pikeminnow, including improved conditions for predation upon juvenile Chinook Salmon ***(add detail here)***. Therefore, predation by Northern Pikeminnow in the Deadwater Slough likely has a consequential impact on ESA-listed Chinook Salmon populations in the Upper Salmon River MPG.

@@

The majority of fish collected during this study received gastric lavage, including some non-predatory species. To validate the efficacy of this method, we euthanized nine Northern Pikeminnow after gastric lavage was completed and removed the remaining stomach contents via dissection. Our results support previous findings that gastric lavage effectively removes stomach contents (@Lott2020). Additionally, we found that the fish captured in net-traps had a similar proportion of stomach content samples. Of those, 1,214 (76.2%) were completely empty, 345 (21.6%) had stomach contents, and 35 (2.2%) contained fish remnants. We were successfully able to identify Redside Shiners, a Largescale Sucker, sculpin, Mountain Whitefish, and one juvenile Chinook Salmon in the stomach contents. However, decomposition from digestion rendered most stomach contents unidentifiable.

@@

The authors extend much appreciation to the many volunteers who assisted with field efforts including collaborators from Bureau of Reclamation, Idaho Department of Fish and Game, and Lemhi Regional Land Trust, among others. This manuscript benefited from reviews and contributions from colleagues at the Idaho Governor's Office of Species Conservation, Rio Applied Science and Engineering, and from Sean Gibbs and Ben Briscoe at Mount Hood Environmental. Funding for this study was provided by the Bureau of Reclamation, Pacific Northwest Regional Office (contract No. 140R1021F0018). Special thanks to Caitlin Alcott and Inter-Fluve for their administrative support and guidance.
